VFTS682: a confirmed dynamical ejection?

### 2018-05-10 VFTS682.tex
fixed sec. 3.1 - numbers are ok now

\section{Introduction}
\label{sec:intro}
\todo{go to major star formation immediately}
\todo{sentence 2:debate of isolated or not}
\todo{paragraph 2 about }
How do massive stars form is one longstanding question in astrophysics
\cite{lada:03, zinnecker:07}. This because they are intrinsically rare
\citep[e.g.,][]{salpeter:55,kroupa:01, schneider:18}, evolve fast, and
remain enshrouded in their parent cloud during the formation
process. Moreover, observations of young massive stars reveal a
complicated multipliticy structure which requires
explanation \citep[][]{sana:12,sana:17}. Understanding massive star formation, possibly as a
function of metallicity, is a key question given the present and upcoming
transient survey \citep[e.g., LSST, BlackGem, LIGO/Virgo O3][]{}\todo{ref} which
will reveal transients associated to massive stars
evolution and death.

Two competing classes of models for massive star formation exist, one predicting 
that massive stars can form in relative isolation \todo{ref}, the other that
they always form in cluster and/or associations, and therefore massive
stars should be rarely isolated \todo{ref}.

The second data release (DR2) from the Gaia satellite
\cite[][]{gaia:16,brown:18} allows us to test
these hypothesis using one particular star, VFTS682. 
This star is a
very massive \citep[$M_\mathrm{ZAMS}\simeq150\,M_\odot$,][]{bestenlehner:11,schneider:18} WNh5 star in the 30 Doradus region
of the Large Magellanic Cloud (LMC), and it is presently observed at a
projected distance of $\sim$$29$\,pc from the nearest cluster of
massive stars R136 \citep[][]{bestenlehner:11}. Based on the extremely high mass of this star and
its present day apparent isolation, \cite{bestenlehner:11} proposed it
might be a candidate for isolated star formation, or a ``slow runaway'' ejected
from R136 in the past. This second option is also supported by the
N-body simulations of \cite{fujii:11, banerjee:12}.   Many other
very massive stars are present in the surroundings of R136, and a more
detailed analysis on the larger sample is desirable.
\todo{this paper raised the debate/interest on VFTS682}

\todo{vfts682 is interesting also because}

Massive stars can in principle be ejected from R136 as a consequence of dynamical
interactions \citep[][]{poveda:67,leonard:91, evans:10, fujii:11,
  allison:12, oh:16}, or by the disruption of a binary by the first
core-collapse supernova
\citep[][]{zwicky:57,blaauw:61,dedonder:97,eldridge:11,renzo:18}. However,
R136 has an estimated age of $\lesssim2$\,Myr \citep[][]{sabbi:12}, which is
shorter than the shortest stellar lifetime
\citep[$\sim$3\,Myr, e.g.,][]{zapartas:17}, so one would not expect the binary
disruption scenario to be relevant for this cluster. \todo{move up}

In this study, we combine the radial velocity measurements from the
VFTS survey \citep[][]{evans:11} with the proper motion from Gaia DR2
to reconstruct the three-dimensional velocity of VFTS682, and test the
hypothesis that this star was ejected from R136. We discuss in
\Secref{sec:sample} the data for VFTS682, and the selection of stars
used to define a local reference frame. Our results
indicate that R136 is a bona fide runaway star (\Secref{sec:runaway}), therefore isolated star formation is \emph{not}
required to explain it. We also find that a dynamical ejection from
R136 is compatible with the direction of its velocity
vector. \todo{double check the following}
However, we find a mild discrepancy between the apparent age of
VFTS682, its kinematic age, and the age estimates for R136 (\Secref{sec:r136_origin}).  We conclude
with a very brief discussion on the implications for theories of
star formation, N-body interactions, and binary evolution in
\Secref{sec:discussion}.

@@

To compare the astrometry of VFTS682 and derive its peculiar motion,
we then select data from the Gaia DR2 catalog for two regions: the
``surroundings'' of VFTS 682, and the ``R136 cluster''. The
surrounding region is defined by all the stars in a target of 10 arcminutes around
VFTS682 fulfilling the following criteria: we require \texttt{visibility\_period} $\geq$ 5,
\texttt{astrometric\_excess\_noise} < 1, the error on the proper
motion components to be smaller than 0.1\,$\mathrm{mas\ yr^{-1}}$,
and the proper motion components themselves to be smaller than
2\,$\mathrm{mas\ yr^{-1}}$ in absolute value. At the distance to the
LMC, $1\mathrm{mas\ yr^{-1}}\simeq250\,\kms$ \citep[e.g.,][]{lennon:18}, so the cut on the values
of the proper motions removes stars that would have projected
tangential velocities in excess of $\sim$500\,$\kms$, which are most
likely to be foreground stars. We checked that the additional
requirement of having parallaxes smaller than $1\,\mathrm{mas\
  yr^{-1}}$ does not reduces further our sample. This selection yields 437 stars.

The ``R136 cluster'' is effectively defined by taking all the stars
in a region of 50 arcseconds around R136a, one of the most massive members of
the cluster itself \citep[][]{crowther:10}, requiring the same
``quality'' criteria applied above. This selection yields 42 stars.
\todo{improve definition of stars from R136, describe accordingly -- check with Danny}

Throughout this study, we assume the same distance of $50$\,kpc to the star, and to
the 30 Doradus region as a whole.% , since the parallax for VFTS682
% listed in the Gaia DR2 catalog is negative. %% this is what makes the
%                                 % LOS velocity useless...how to improve?

For each of the two local rest frames (``surroundings'' and ``R136
cluster''), we compute the average transverse velocity as:

\begin{equation}
  \label{eq:mean}
  \langle \mu_i\rangle = \frac{\sum_\mathrm{stars}\frac{1}{\Delta
      \mu_i}\mu_i}{\sum_\mathrm{stars} \frac{1}{\Delta \mu_i}} \ \ , \
  \ \Delta \langle \mu_i\rangle = \frac{\sqrt{N}}{\sum_\mathrm{stars}
    \frac{1}{\Delta \mu_i}} \ \ ,
\end{equation}
where $i = \mathrm{RA}, \mathrm{DEC}$, and $\Delta \mu_i$ is the error
on the proper motion component reported by Gaia, and the sums run over
all the $N$ stars of the frame considered. We evaluate each proper motion
component separately.

The data retrieved, and the ipython notebook used for the analysis
presented here will be made available at \todo{probably git repo on bitbucket?}.

@@

Using the 437 stars selected as the ``surrounding population'' in
\Secref{sec:sample} and shown in blue in \Figref{fig:main}, we find averaged proper motion components of
$\langle\mu_\mathrm{RA}^\mathrm{sur}\rangle = 1.695\pm0.003\,\mathrm{mas\ yr^{-1}}$ and
$\langle\mu_\mathrm{DE}^\mathrm{sur}\rangle = 0.691\pm0.003\,\mathrm{mas\ yr^{-1}}$. Subtracting these values from the
proper motions of VFTS682 (see \Tabref{tab:vfts682}), we obtain the
components of proper motion of the star relative to the surrounding region
$\mu^\mathrm{sur}_\mathrm{RA} = 0.15\pm 0.07\,\mathrm{mas\ yr^{-1}}$ and $\mu^\mathrm{sur}_\mathrm{DE} =
0.09\pm 0.08\,\mathrm{mas\ yr^{-1}}$. %% note that the error budget is
% dominated by VFTS682
These can be converted in the
the components of the transverse velocity $v^\mathrm{sur}_\mathrm{RA}=35\pm17\,\kms$,
$v^\mathrm{sur}_\mathrm{DE}=22\pm19\,\kms$, assuming a distance of
50\,kpc (we do not account for the uncertainty in the distance
estimate when propagating errors). The radial velocity from
\cite{bestenlehner:11} then gives the third component along
the line of sight, allowing us to calculate the speed of the star:

\begin{equation}
  \label{eq:speed_around}
  v^\mathrm{sur} = \sqrt{\left(v^\mathrm{sur}_\mathrm{RA}\right)^2
    +\left(v^\mathrm{sur}_\mathrm{DE}\right)^2+\left(\delta v_\mathrm{rad}\right)^2} = 61 \pm 22
  \, \kms \ .
\end{equation}
This value for the three-dimensional speed of VFTS682 with respect the
surrounding stars make it a ``bona fide'' runaway star.

@@

The inset plot in \Figref{fig:main} shows a zoom in of the region
around R136 which includes VFTS682. Stars belonging to our ``R136
cluster'' group are shown in green (see \Secref{sec:sample}), together with their proper motion
as listed in the Gaia DR2 catalog.
To test the hypothesis that VFTS682 was indeed ejected from this region, we check the
orientation of its proper motion relative to these stars. We further
consider also the kinematic age of the star.

We begin by evaluating the average proper motion for the stars
representing the cluster. This is necessary to constrain the actual velocity
obtained by VFTS682 at its ejection. By using \Eqref{eq:mean} we find  
$\langle\mu_\mathrm{RA}^{R136}\rangle = 1.750\pm0.008\,\mathrm{mas\ yr^{-1}}$
and $\langle\mu_\mathrm{DE}^\mathrm{R136}\rangle =
0.742\pm0.009\,\mathrm{mas\ yr^{-1}}$. Therefore, using the values
from \Tabref{tab:vfts682}, we obtain the components of the proper
motion for VFTS682 relative to R136 $\mu^{R136}_\mathrm{RA} = 0.09\pm 0.07\,\mathrm{mas\ yr^{-1}}$ and $\mu^{R136}_\mathrm{DE} =
0.04\pm 0.08\,\mathrm{mas\ yr^{-1}}$.\todo{improve numbers by
  selecting better sample}.
\todo{discuss if direction is compatible
The error cone, obtained by
propagating the errors on the proper motion components indicates that
R136 is indeed the most likely origin, as expected by \cite{fujii:11,
  banerjee:12}.
}

Assuming again a distance of 50\,kpc and
not considering the error on the distance, we can convert these into
the components of the tangential velocity relative to R136,
$v^\mathrm{R136}_\mathrm{RA}=21 \pm 18\, \kms$ and $v^\mathrm{R136}_\mathrm{DE}=10
\pm 21\,\kms$. Finally, using the peculiar line of sight velocity
from \cite{bestenlehner:11}, we obtain the three-dimensional peculiar
velocity of VFTS682 with respect to R136

\begin{equation}
  \label{eq:speed_R136}
  v^\mathrm{sur} = \sqrt{\left(v^{R136}_\mathrm{RA}\right)^2
    +\left(v^{R136}_\mathrm{DE}\right)^2+\left(\delta v_\mathrm{rad}\right)^2} = 51 \pm 23
  \, \kms \ .
\end{equation}

\todo{number above not trustworthy}

\todo{kinematic age}

@@

\section{Summary and Discussion}
\label{sec:discussion}

\todo{compare to Lennon: time of ejection is the same? what does it
  say for cluster}
\todo{say implications for center of R136 (same stars are there), how
  many ejected given this one, was a binary ejected in the other direction?}
\todo{does this imply there is a binary in the core ?}
\begin{itemize}
\item VFTS682 is a bona fide runaway with $v\sim60\,\mathrm{km\
    s^{-1}}$ thrown out from R136. Both its speed and the age of the
  cluster are consitent with a dynamical ejection.
  \item VFTS682 comes from R136 as was expected by
  \cite{bestenlehner:11, fujii:11, banerjee:12}, so it does not
  require isolated SFH to be explained
\item apparent age tension (connect to VFTS16 as well).
\item is R136 a single young cluster or a merger
\item estimate the influence of the gravitational potential of R136,
  what is its total mass and relaxation time?
\end{itemize}

@@

>>>>

\section{Introduction}
\label{sec:intro}

How do massive stars form is one longstanding question in astrophysics
\citep[e.g.,][]{lada:03, zinnecker:07}, because they are intrinsically rare
\citep[e.g.,][]{salpeter:55,kroupa:01, schneider:18}, evolve fast, and
remain enshrouded in their parent cloud during the formation
process. Moreover, observations of young massive stars reveal a
complicated multipliticy structure which requires
explanation \citep[][]{sana:12,sana:17}. Understanding massive star formation, possibly as a
function of metallicity, is a key question given the present and upcoming
transient survey \citep[e.g., LSST, BlackGem, LIGO/Virgo O3][]{}\todo{ref} which
will reveal transients associated to massive stars
evolution and death.

Two competing classes of models for massive star formation exist. One class predicts 
that massive stars can form in relative isolation \todo{ref}, while the other predicts
they always form in cluster and/or associations, and therefore
isolated massive stars should necessarily peculiar velocities to reach
an isolated position on the sky\todo{ref}.

The second data release (DR2) from the Gaia satellite
\cite[][]{gaia:16,brown:18} allows us to test
these hypothesis using one particular star, VFTS682. 
This star belongs to the 30 Doradus region in the Large Magellanic
Cloud (LMC), and has an inferred initial mass of $M_\mathrm{ZAMS}=150.0^{+28.7}_{-17.4}\,M_\odot$
\citep[][]{bestenlehner:11,schneider:18}. Its spectral type is WNh5, and it is presently observed at a
projected distance of $\sim$$29$\,pc from the nearest cluster of
massive stars R136 \citep[also known as NGC2070][]{bestenlehner:11}. The inferred mass loss
rate is $\sim10^{-XX}\,M_\odot \ \mathrm{yr}^{-1}$ \todo{ref}.

Based on the extremely high mass of this star and
its present-day apparent isolation, \cite{bestenlehner:11} proposed it
could be a candidate for isolated star formation, or a ``slow runaway'' ejected
from R136 in the past. Their study raised the interest in the origin
of this star. \cite{fujii:11, banerjee:12} carried out N-body
simulations of massive young stellar clusters and concluded that the
ejection of such a massive object from R136 through dynamical
interactions \citep[e,g,][]{poveda:67} would be possible.
The competing ejection mechanism, the disruption of a binary
system by a core-collapse event \citep[][]{zwicky:57, blaauw:61}, is
thought to be ruled out, since the cluster has an estimated age of
$\lesssim2$\,Myr \citep[][]{sabbi:12}, which is shorter than the shortest stellar lifetime
\citep[$\sim$3\,Myr, e.g.,][]{zapartas:17}.

In this study, we combine the radial velocity measurements from the
VFTS survey \citep[][]{evans:11} with the proper motion from Gaia DR2
to reconstruct the three-dimensional velocity of VFTS682, and test the
hypothesis that this star was ejected from R136. We discuss in
\Secref{sec:sample} the data for VFTS682, and the selection of stars
used to define a local reference frame. Our results
indicate that R136 is a bona fide runaway star (\Secref{sec:runaway}),
therefore isolated star formation is \emph{not}
required to explain it. We also find that a dynamical ejection from
R136 is compatible with the direction of its velocity
vector. 
We conclude
with a very brief discussion on the implications for theories of
star formation, N-body interactions, and binary evolution in
\Secref{sec:discussion}.

@@

We define a local standard frame to derive the peculiar velocity
of VFTS682 by selecting from the Gaia DR2 catalog a sample of nearby
stars, following closely the approach of \cite{lennon:18}.
We select all the stars in a target of 0.2 degrees around R136
(NGC2070) fulfilling the following criteria. First, we require G-band
magnitude brighter than 17, correspondingly roughly to the
completeness level of the VFTS survey \citep[here we implicitly assume
G$\sim$V,][]{evans:11}. Then we require \texttt{visibility\_period} $\geq$ 5,
\texttt{astrometric\_excess\_noise} < 1, the errors on the proper
motion components to be smaller than 0.1\,$\mathrm{mas\ yr^{-1}}$,
and the proper motion components themselves to be smaller than
2\,$\mathrm{mas\ yr^{-1}}$ in absolute value. At the distance to the
LMC, $1\mathrm{mas\ yr^{-1}}\simeq250\,\kms$ \citep[e.g.,][]{lennon:18}, so the cut on the values
of the proper motions removes stars that would have projected
tangential velocities in excess of $\sim$500\,$\kms$, which are most
likely to be foreground stars. We checked that the additional
requirement of having parallaxes smaller than $2\,\mathrm{mas}$ does
not reduces further our sample.

We calculate the averaged proper motion components for the whole
region using 
\begin{equation}
  \label{eq:mean}
  \langle \mu_i\rangle = \frac{\sum_\mathrm{stars}\frac{1}{\Delta
      \mu_i}\mu_i}{\sum_\mathrm{stars} \frac{1}{\Delta \mu_i}} \ \ , \
  \ \Delta \langle \mu_i\rangle = \frac{\sqrt{N}}{\sum_\mathrm{stars}
    \frac{1}{\Delta \mu_i}} \ \ ,
\end{equation}
where $i = \mathrm{RA}, \mathrm{DEC}$, and $\Delta \mu_i$ is the error
on the proper motion component reported by Gaia. The sums run over
all the $N=651$ stars in our selected sample. We evaluate each proper motion
component separately. For simplicity, throughout this study, we assume the same
distance of $50$\,kpc to the star \citep[][]{lebouteiller:08}, and to
the 30 Doradus region as a whole. We do not consider the error bars on
the distance determination when converting proper motions into
physical velocities. The data retrieved, and the ipython notebook used for the analysis
presented here will be made available at \todo{probably git repo on bitbucket?}.

@@

Using the 651 stars selected as described in \Secref{sec:sample} and shown in blue in \Figref{fig:main}, we find averaged proper motion components of
$\langle\mu_\mathrm{RA}^\mathrm{sur}\rangle = 1.683\pm0.002\,\mathrm{mas\ yr^{-1}}$ and
$\langle\mu_\mathrm{DE}^\mathrm{sur}\rangle =
0.672\pm0.003\,\mathrm{mas\ yr^{-1}}$. We note that these values are
in good agreement with what found by \cite{lennon:18}. Subtracting these values from the
proper motions of VFTS682 (see \Tabref{tab:vfts682}), we obtain the
components of proper motion of the star relative to the surrounding region
$\mu^\mathrm{sur}_\mathrm{RA} = 0.16\pm 0.07\,\mathrm{mas\ yr^{-1}}$ and $\mu^\mathrm{sur}_\mathrm{DE} =
0.11\pm 0.08\,\mathrm{mas\ yr^{-1}}$.
These can be converted in the
the components of the transverse velocity $v^\mathrm{sur}_\mathrm{RA}=38\pm17\,\kms$,
$v^\mathrm{sur}_\mathrm{DE}=26\pm19\,\kms$, assuming a distance of
50\,kpc (we do not account for the uncertainty in the distance
estimate when propagating errors). The radial velocity from
\cite{bestenlehner:11} then gives the third component along
the line of sight, allowing us to calculate the speed of the star:

\begin{equation}
  \label{eq:speed_around}
  v^\mathrm{sur} = \sqrt{\left(v^\mathrm{sur}_\mathrm{RA}\right)^2
    +\left(v^\mathrm{sur}_\mathrm{DE}\right)^2+\left(\delta
      v_\mathrm{rad}\right)^2} = 64 \pm 21 
  \, \kms \ .
\end{equation}
This value for the three-dimensional speed of VFTS682 with respect the
surrounding stars make it the most massive ``bona fide'' runaway star
known to date.

@@

% The inset plot in \Figref{fig:main} shows a zoom in of the region
% around R136 which includes VFTS682. Stars belonging to our ``R136
% cluster'' group are shown in green (see \Secref{sec:sample}), together with their proper motion
% as listed in the Gaia DR2 catalog.
% To test the hypothesis that VFTS682 was indeed ejected from this region, we check the
% orientation of its proper motion relative to these stars. We further
% consider also the kinematic age of the star.

% We begin by evaluating the average proper motion for the stars
% representing the cluster. This is necessary to constrain the actual velocity
% obtained by VFTS682 at its ejection. By using \Eqref{eq:mean} we find  
% $\langle\mu_\mathrm{RA}^{R136}\rangle = 1.750\pm0.008\,\mathrm{mas\ yr^{-1}}$
% and $\langle\mu_\mathrm{DE}^\mathrm{R136}\rangle =
% 0.742\pm0.009\,\mathrm{mas\ yr^{-1}}$. Therefore, using the values
% from \Tabref{tab:vfts682}, we obtain the components of the proper
% motion for VFTS682 relative to R136 $\mu^{R136}_\mathrm{RA} = 0.09\pm 0.07\,\mathrm{mas\ yr^{-1}}$ and $\mu^{R136}_\mathrm{DE} =
% 0.04\pm 0.08\,\mathrm{mas\ yr^{-1}}$.\todo{improve numbers by
%   selecting better sample}.
% \todo{discuss if direction is compatible
% The error cone, obtained by
% propagating the errors on the proper motion components indicates that
% R136 is indeed the most likely origin, as expected by \cite{fujii:11,
%   banerjee:12}.
% }

% Assuming again a distance of 50\,kpc and
% not considering the error on the distance, we can convert these into
% the components of the tangential velocity relative to R136,
% $v^\mathrm{R136}_\mathrm{RA}=21 \pm 18\, \kms$ and $v^\mathrm{R136}_\mathrm{DE}=10
% \pm 21\,\kms$. Finally, using the peculiar line of sight velocity
% from \cite{bestenlehner:11}, we obtain the three-dimensional peculiar
% velocity of VFTS682 with respect to R136

% \begin{equation}
%   \label{eq:speed_R136}
%   v^\mathrm{sur} = \sqrt{\left(v^{R136}_\mathrm{RA}\right)^2
%     +\left(v^{R136}_\mathrm{DE}\right)^2+\left(\delta v_\mathrm{rad}\right)^2} = 51 \pm 23
%   \, \kms \ .
% \end{equation}

\subsection{Kinematic age}

@@

\section{Discussion}
\label{sec:discussion}

\todo{compare to Lennon: time of ejection is the same? what does it
  say for cluster}
\todo{say implications for center of R136 (same stars are there), how
  many ejected given this one, was a binary ejected in the other direction?}
\todo{does this imply there is a binary in the core ?}
\begin{itemize}
\item VFTS682 is a bona fide runaway with $v\sim60\,\mathrm{km\
    s^{-1}}$ thrown out from R136. Both its speed and the age of the
  cluster are consitent with a dynamical ejection.
  \item VFTS682 comes from R136 as was expected by
  \cite{bestenlehner:11, fujii:11, banerjee:12}, so it does not
  require isolated SFH to be explained
\item apparent age tension (connect to VFTS16 as well).
\item is R136 a single young cluster or a merger
\item estimate the influence of the gravitational potential of R136,
  what is its total mass and relaxation time?
\end{itemize}
\section{Summary}

1. Gaia data confirm a 150Msun runaway with v~60, thus the most
massive known to date
2. directions suggest ejection from R136, confirming prediction of
fuji+ banerjee+
2b. This fits with the observation from crowther of M>300Msun in R136
(and mention vfts682 has the same spectral type): the most likely
mechanism needs encounter with 2 more massive -> additional evidence
for Crowther claim
3. comment on final fate (BH)

@@

Many other very massive stars are present in the surroundings this cluster, and a more
detailed analysis on the larger sample is desirable.
